### 2015-04-06 cug_paper/01_introduction.tex
Scott's intro edits

In most HPC applications, filesystem writes typically follow a very `bursty' pattern. That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.  Also, at ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it. This means that the performance that a user sees at any given time will likely vary dramatically.

Having a sufficiently large node-local write-back cache could improve applications' performance in two ways.  First, if the cache is large enough to hold an entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.  It can write to the cache and then resume its computations while the cache drains to the filesystem.  Secondly, write performance should be more predictable. From the application's point of view, write speed would only limited by how fast data can be copied into the cache, and since the cache is local to the node, it is not effected by outside influences.

Titan, the Cray XK7 at Oak Ridge National Laboratory, has 18,688 compute nodes. Each node consists of a 16-core AMD CPU, an NVIDIA GPU and 32GB ram. In addition, there is another 6GB of ram on each GPU card. Not surprisingly, there are applications that do not use all the resources on a compute node.

To start with, during 2014, approximately 50\% of the compute time on Titan was used on applications that did not use the GPU at all.\cite{oareport}
There may be good reasons for this.  Utilizing the GPU's requires changes to the application and it's possible the developers haven't had the time to do that.  It's also possible that the type of computations the application performs simply don't lend themselves to GPU-based acceleration.  Regardless of why, it's clear that not all applications are benefiting from the GPU.

Secondly, though AMD markets the processor as having 16 cores, it actually has only 8 floating-point units. As such, some applications that are floating-point intensive run with only 8 processes per node. This actually leaves 8 cores available so long as they primarily execute integer instructions.

 The unused GPU memory and integer cores can be used to implement a write-back cache without taking resources away from the main application.  This paper demonstrates a technique for providing such a cache to applications.   Runs were made on Titan using a synthetic benchmark to demonstrate performance both without such a cache and with a cache that utilizes ram on the GPU card. The results show definite performance benefits.

\begin{comment}

>>>>

Most HPC applications' I/O exhibit a very `bursty' pattern. That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.  At ORNL, the filesystem is a center-wide resource shared with other systems including a Cray XC-30, analysis clusters, and WAN-connected data mover nodes. Each of these competing systems place a varying load on the filesystem such that Titan's users' write performance will vary dramatically within a job and between jobs.

Having a sufficiently large, node-local, write-back cache could improve an application's performance in two ways.  First, if the cache is large enough to hold an entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.  It can write to the cache and then resume its computations while the cache drains to the filesystem.  Secondly, write performance should be more predictable. From the application's point of view, write speed would only limited by how fast data can be copied into the cache, and since the cache is local to the node, it is not effected by outside influences.

Titan, the Cray XK7 at Oak Ridge National Laboratory, has 18,688 compute nodes. Each node consists of a 16-core AMD Interlagos CPU, an NVIDIA GPU and 32GB DRAM. In addition, there is another 6GB of GDDR DRAM on each GPU card. Not surprisingly, some applications do not use all the resources on a compute node.

To start with, during 2014, approximately 50\% of the compute time on Titan was used on applications that did not use the GPU at all.\cite{oareport}
There may be good reasons for this.  Utilizing the GPU requires changes to the application and it is possible the developers have not had the time or funding to do that.  It is also possible that the type of computations the application performs simply do not lend themselves to GPU-based acceleration (i.e. not enough exposed parallelism).  Regardless of why, it is clear that not all applications are benefiting from the GPU.

Secondly, the Interlagos has eight Bulldozer units, each of which has two full integer units and a shared floating-point unit. Some users with floating-point intensive applications run jobs using only eight processes per node (one process per Bulldozer) to avoid oversubscribing the floating-point units. This actually leaves eight integer cores available so long as they primarily execute integer instructions.

The unused GPU memory and integer cores can be used to implement a write-back cache without taking resources away from the main application.  This paper demonstrates a technique for providing such a cache to applications.   Runs were made on Titan using a synthetic benchmark to demonstrate performance both without such a cache and with a cache that utilizes DRAM on the GPU card. The results show definite performance benefits.

%
% Everything below is commented out.  It's text I've toyed with using, but decided not to for whatever reason
%
\begin{comment}
